Type Error Telemetry at Scale

### 2023-09-27 programming23/paper.tex
related work ++
sec5 threats

Creators build experiences using the
{Luau} programming language~\cite{luau-lang.org},
an extension of {Lua~5.1~\cite{lua}}.
The main addition in Luau is a static type system, which infers
types for all Luau programs on the fly, as creators modify the code.
These types are used primarily in type-driven tooling such as autocomplete and
API documentation~\cite{luau-autocomplete}; the typical creator is unaware
that a type system powers these tools.
However, creators may write their own types, and they can opt in
to receiving ype error reports.

@@

With this telemetry data, we investigate research questions about
the adoption and benefits of type analysis:
\begin{description}
  \item[RQ1.]
    How many creators use type analysis?
    How often do projects contain modules with different
    analysis modes?
    How often do creators turn analysis off?
  \item[RQ2.]
    For modules that use type analysis:
    which errors arise,
    how do creators respond,
    and which errors tend to persist despite subsequent edits?
  \item[RQ3.]
    What impact does type analysis have on the number of background type
    errors?
    For example, do background errors pile up in unanalyzed projects?
\end{description}

@@

FILL can stats answer questions that point-by-point analysis cannot?

@@

  \begin{tabular}{l}
    600K type errors         = \pct{35} \mnocheck{}, \pct{61} \mnonstrict{}, \pct{4} \mstrict{}
    \\
    72M forced scrict errors = \pct{89} \mnocheck{}, \pct{10} \mnonstrict{}, \pct{1} \mstrict{}
    %% really 0.16 strict, very low!!!
  \end{tabular}
  \caption{}
  \label{f:error-by-mode}
\end{figure}

@@

%% \paragraph{Most Type Errors are Typos}
\begin{itemize}
  \item
    One striking aspect of~\cref{t:type-error-survival} is the large number
    of \code{SyntaxError}s and the related errors \code{UnknownSymbol} and
    \code{UnknownProperty}.
    Most type errors in the data are likely due to typos.
    \Cref{s:type-error-count} discusses this point in detail.
    For now, we remark that \code{SyntaxError} is the only possible error
    in \mnocheck{} mode (by design) and observe that these errors are
    usually in the Drop column and seldom in the Keep column.\todo{So?}

@@

\begin{table}[t]\centering
  %% NOTE: popularity looks similar when [over all records] vs. [restricted to module switch records]
  \caption{Type error popularity for \mnonstrict{} and \mstrict{} modes.  All of the type errors in \mnocheck{} mode are \code{SyntaxError}s.}
  \label{t:type-error-count}
  \begin{tabular}{ll}
    \mnonstrict{} & \mstrict{} \\
    \begin{tabular}[t]{l@{~}r}
      \code{UnknownSymbol} & \pct{62.13} \\
      \code{SyntaxError} & \pct{15.42} \\
      \code{UnknownProperty} & \pct{8.28} \\
      \code{UnknownRequire} & \pct{3.13} \\
      \code{TypeMismatch} & \pct{2.44} \\
      \code{CountMismatch} & \pct{2.26} \\
      \code{OptionalValueAccess} & \pct{2.07} \\
      \code{GenericError} & \pct{2.06} \\
      \code{UnknownPropButGotLikeProp} & \pct{0.44} \\
      \code{CannotExtendTable} & \pct{0.42} \\
      \code{GenericExtraInformation} & \pct{0.32} \\
      \code{ModuleHasCyclicDependency} & \pct{0.23} \\
      \code{IllegalRequire} & \pct{0.20} \\
      \code{NotATable} & \pct{0.15} \\
      \code{CannotCallNonFunction} & \pct{0.15} \\
      \code{MissingProperties} & \pct{0.09} \\
      \code{FunctionExitsWithoutReturn} & \pct{0.07} \\
      \code{FunctionDoesNotTakeSelf} & \pct{0.07} \\
      \code{MissingUnionProperty} & \pct{0.02} \\
      \code{CannotInferBinaryOperation} & \pct{0.02} \\
      \code{OnlyTablesCanHaveMethods} & \pct{0.01} \\
      \code{DuplicateTypeDefinition} & {<\pct{0.01}} \\
      \code{TypesAreUnrelated} & {<\pct{0.01}} \\
    \end{tabular}
        &
    \begin{tabular}[t]{l@{~}r}
      \code{UnknownSymbol} & \pct{23.97} \\
      \code{TypeMismatch} & \pct{20.46} \\
      \code{UnknownProperty} & \pct{18.88} \\
      \code{SyntaxError} & \pct{9.31} \\
      \code{CannotInferBinaryOperation} & \pct{6.94} \\
      \code{MissingProperties} & \pct{4.04} \\
      \code{CountMismatch} & \pct{2.99} \\
      \code{OptionalValueAccess} & \pct{2.99} \\
      \code{FunctionExitsWithoutReturn} & \pct{2.99} \\
      \code{UnknownRequire} & \pct{2.37} \\
      \code{GenericError} & \pct{2.28} \\
      \code{NotATable} & \pct{1.32} \\
      \code{GenericExtraInformation} & \pct{0.35} \\
      \code{UnknownPropButGotLikeProp} & \pct{0.26} \\
      \code{IncorrectGenericParamCount} & \pct{0.18} \\
      \code{CannotCallNonFunction} & \pct{0.18} \\
      \code{IllegalRequire} & \pct{0.18} \\
      \code{ModuleHasCyclicDependency} & \pct{0.09} \\
      \code{CannotExtendTable} & \pct{0.09} \\
      \code{OccursCheckFailed} & \pct{0.09} \\
      \code{TypesAreUnrelated} & \pct{0.09} \\
    \end{tabular}
  \end{tabular}
\end{table}

@@

@@

  \mnocheck{}
  \includegraphics[width=\columnwidth]{img/error-count-nocheck-row--te-density-diff.pdf}
  \medskip
  \mnonstrict{}
  \includegraphics[width=\columnwidth]{img/error-count-nonstrict-row--te-density-diff.pdf}
  \medskip
  \mstrict{}
  \includegraphics[width=\columnwidth]{img/error-count-strict-row--te-density-diff.pdf}
  \caption{Changes in type error density (errors in module / lines in module) over time (seconds).}
  \label{f:error-density}
\end{figure}

@@

\section{Discussion}
\label{s:discussion}

To sum up, the data supports the following conclusions:

\subsection{The Good News}

\paragraph{Good: No Internal Error}

FILL hope to never see \code{InternalError}.
Have zero, yay!

um ... very concerning if this is the top takeaway

\paragraph{Good: No Apparent Turmoil With Upgrades / Downgrades}

But, mode switches are very rare!
Really too soon to draw any conclusion, this is only a lack of bad.

\paragraph{Do Types Improve Quality of Life?}

\FILL{} compare \mnonstrict{} and \mstrict{}.
Use edit range~(\cref{t:type-error-survival} for NS vs S)
and error density~(\cref{f:error-density})

Uh oh, no accumulated FS errors in nocheck.
Edits vs FS counts are the same in all three modes.

\paragraph{... Something About Add/Drop in Edit Range}

\FILL{}

\subsection{The Bad News}

\paragraph{Bad: Strict too Strict}

\FILL{}

Comparing NS vs S, the syntax error \% is about the same (15 vs 9)
but the type mismatch \% is far off (2 vs 20).
S is probably noisy.

Data model is probably part of this.

\paragraph{Big Bad: Low Adoption}

poor advertising,
unclear value proposition,
nocheck code is apparently doing fine despite lots of forced strict errors,
... can we say this outright?!

\subsection{Threats to Validity}
\label{s:threats}

Error counts per-project and per-module include syntax errors.
Problematic, because syntax errors are uninteresting (easy to fix),
but common, which makes it likely that our per-keystroke telemetry (1/200) will record them.
Edit range has specific counts.
Next time, we should do that for the other aggregates.

We are sampling sessions. Could miss interesting ones.

Sampling by keystroke means different things for different creators.
Depends how they type.
Scripted interactions that generate a large number of keystrokes per second
can swamp our data and are hard to distinguish from genuine interactions.
Copy paste might just look like a bot.
%% \QALAN{} I think you said a paste counts as 1 keystroke; is that right?

We have no knowledge whatsoever about interesting events like save run publish.
Publish events are private.

Changes to errors in the edit range are hard to follow.
Could add one error but remove another of the same kind,
then we can't distinguish from ignoring the error.

Edit ranges are coarse, merely the top and bottom edit.
No idea what happened in between.
Long time between keystroke events makes this even worse.

Telemetry cannot replace user studies, as there is no way to measure
creator sentiment, but provide complementary data at scale.
No idea about intent is a big problem, we don't know which error the creator wanted to fix.

Type analysis reports many errors at once.
No way to know what the creator is looking at.
Not even sure they have the analysis widget open.

Internal validity?

External validity?

Construct validity?

Conclusion validity

Ecological validity is high.

@@

\begin{table}[t]\centering
  \caption{Comparison of telemetry designs. Two orthogonal concerns are when to collect data and whether to add noise to it.}
  \label{t:telemetry-design}

@@

The Blackbox project similarly collects telemetry in broad strokes.
Every Java program submitted for compilation goes into the Blackbox database (minus
any comments above the main class) along with the compiler's response~\cite{bkmu-sigcse-2014}.
%% Though somewhat intrusive,
The millions of programs in the Blackbox dataset
have enabled dozens of contributions to the area of CS
education~\cite{bask-icer-2018}, ranging from categorization of novice
errors~\cite{mk-fie-2014,m-masters-2016} to analyses of error frequency and
time-to-fix~\cite{ab-sigcse-2015}.

@@

%\paragraph{TBD}
%
%thomas kennedy .... generic usage monitoring ascilite 2003
%
%ahmadzadeh elliman higgins iticse 2005
%
%buffardi etal 2014 adaptive and social mechanisms
%
%bddf-icse-2016
% sh edwards and jason snyder icer 2009 comparing effective and ineffective grading platform
%
%murphy etal sigcse 2009 Retina: helping students
%
%Choppella Haynes IU TR 426

@@

>>>>

Creators build experiences using the
{Luau} programming language~\cite{luau-lang.org},
an extension of {Lua~5.1~\cite{lua}}.
The main addition in Luau is a static type system, which infers
types for all Luau programs on the fly, as creators modify the code.
These types are used primarily in type-driven tooling such as autocomplete and
API documentation~\cite{luau-autocomplete}; the typical creator is unaware
that a type system powers these tools.
However, creators may write their own types, and they can opt in
to receiving type error reports.

@@

With this telemetry data, we investigate research questions about
the adoption and benefits of type analysis:
\begin{description}
  \item[RQ1.]
    How many creators use type analysis?
    How often do projects contain modules with different
    analysis modes?
    How often do creators turn analysis off?
  \item[RQ2.]
    For modules that use type analysis:
    which errors arise,
    how do creators respond,
    and which errors tend to persist despite subsequent edits?
  \item[RQ3.]
    What impact does type analysis have on the number of \FS{}
    errors?
    For example, do \FS{} errors pile up in unanalyzed projects?
\end{description}

@@

@@

  \begin{tabular}{l}
    600K type errors         = \pct{35} \mnocheck{}, \pct{61} \mnonstrict{}, \pct{4} \mstrict{}
    \\
    72M forced scrict errors = \pct{89} \mnocheck{}, \pct{10} \mnonstrict{}, \pct{1} \mstrict{}
    %% really 0.16 strict, very low!!!
  \end{tabular}
  \caption{FILL}
  \label{f:error-by-mode}
\end{figure}

@@

%% \paragraph{Most Type Errors are Typos}
\begin{itemize}
  \item
    \FILL{} which survive the longest, most keeps???

  \item
    One striking aspect of~\cref{t:type-error-survival} is the large number
    of \code{SyntaxError}s and the related errors \code{UnknownSymbol} and
    \code{UnknownProperty}.
    Most type errors in the data are likely due to typos.
    \Cref{s:type-error-count} discusses this point in detail.
    For now, we remark that \code{SyntaxError} is the only possible error
    in \mnocheck{} mode (by design) and observe that these errors are
    usually in the Drop column and seldom in the Keep column.\todo{So?}

@@

\begin{table}[t]\centering
  %% NOTE: popularity looks similar when [over all records] vs. [restricted to module switch records]
  \caption{Type error popularity for \mnonstrict{} and \mstrict{} modes (all of the type errors in \mnocheck{} mode are \code{SyntaxError}s).}
  \label{t:type-error-count}
  \begin{tabular}{ll}
    \mnonstrict{} & \mstrict{} \\
    \begin{tabular}[t]{l@{~}r}
      \code{UnknownSymbol} & \pct{62.13} \\
      \code{SyntaxError} & \pct{15.42} \\
      \code{UnknownProperty} & \pct{8.28} \\
      \code{UnknownRequire} & \pct{3.13} \\
      \code{TypeMismatch} & \pct{2.44} \\
      \code{CountMismatch} & \pct{2.26} \\
      \code{OptionalValueAccess} & \pct{2.07} \\
      \code{GenericError} & \pct{2.06} \\
      \code{UnknownPropButGotLikeProp} & \pct{0.44} \\
      \code{CannotExtendTable} & \pct{0.42} \\
      \code{GenericExtraInformation} & \pct{0.32} \\
      \code{ModuleHasCyclicDependency} & \pct{0.23} \\
      \code{IllegalRequire} & \pct{0.20} \\
      \code{NotATable} & \pct{0.15} \\
      \code{CannotCallNonFunction} & \pct{0.15} \\
      \code{MissingProperties} & \pct{0.09} \\
      \code{FunctionExitsWithoutReturn} & \pct{0.07} \\
      \code{FunctionDoesNotTakeSelf} & \pct{0.07} \\
      \code{MissingUnionProperty} & \pct{0.02} \\
      \code{CannotInferBinaryOperation} & \pct{0.02} \\
      \code{OnlyTablesCanHaveMethods} & \pct{0.01} \\
      \code{DuplicateTypeDefinition} & {<\pct{0.01}} \\
      \code{TypesAreUnrelated} & {<\pct{0.01}} \\
    \end{tabular}
        &
    \begin{tabular}[t]{l@{~}r}
      \code{UnknownSymbol} & \pct{23.97} \\
      \code{TypeMismatch} & \pct{20.46} \\
      \code{UnknownProperty} & \pct{18.88} \\
      \code{SyntaxError} & \pct{9.31} \\
      \code{CannotInferBinaryOperation} & \pct{6.94} \\
      \code{MissingProperties} & \pct{4.04} \\
      \code{CountMismatch} & \pct{2.99} \\
      \code{OptionalValueAccess} & \pct{2.99} \\
      \code{FunctionExitsWithoutReturn} & \pct{2.99} \\
      \code{UnknownRequire} & \pct{2.37} \\
      \code{GenericError} & \pct{2.28} \\
      \code{NotATable} & \pct{1.32} \\
      \code{GenericExtraInformation} & \pct{0.35} \\
      \code{UnknownPropButGotLikeProp} & \pct{0.26} \\
      \code{IncorrectGenericParamCount} & \pct{0.18} \\
      \code{CannotCallNonFunction} & \pct{0.18} \\
      \code{IllegalRequire} & \pct{0.18} \\
      \code{ModuleHasCyclicDependency} & \pct{0.09} \\
      \code{CannotExtendTable} & \pct{0.09} \\
      \code{OccursCheckFailed} & \pct{0.09} \\
      \code{TypesAreUnrelated} & \pct{0.09} \\
    \end{tabular}
  \end{tabular}
\end{table}

%% UNUSED CODES
%% \mnonstrict{}
%%   CodeTooComplex
%%   DeprecatedApiUsed
%%   DuplicateGenericParameter
%%   DynamicPropertyLookupOnClassesUnsafe
%%   FunctionRequiresSelf
%%   IncorrectGenericParameterCount
%%   InternalError
%%   NormalizationTooComplex
%%   OccursCheckFailed
%%   SwappedGenericTypeParameter
%%   TypePackMismatch
%%   UnificationTooComplex
%% 
%% \mstrict{}
%%  CodeTooComplex
%%  DeprecatedApiUsed
%%  DuplicateGenericParameter
%%  DuplicateTypeDefinition
%%  DynamicPropertyLookupOnClassesUnsafe
%%  FunctionDoesNotTakeSelf
%%  FunctionRequiresSelf
%%  InternalError
%%  MissingUnionProperty
%%  NormalizationTooComplex
%%  OnlyTablesCanHaveMethods
%%  SwappedGenericTypeParameter
%%  TypePackMismatch
%%  UnificationTooComplex

%% both NS S
%%  CodeTooComplex
%%  DeprecatedApiUsed
%%  DuplicateGenericParameter
%%  DynamicPropertyLookupOnClassesUnsafe
%%  FunctionRequiresSelf
%%  InternalError
%%  NormalizationTooComplex
%%  OnlyTablesCanHaveMethods
%%  SwappedGenericTypeParameter
%%  TypePackMismatch
%%  UnificationTooComplex

@@

  \item
    \FILL{} huge syntax error gap between S nonstrict.

@@

  \mnocheck{}
  \includegraphics[width=\columnwidth]{img/error-count-nocheck-row--te-density-diff.pdf}
  \smallskip

  \mnonstrict{}
  \includegraphics[width=\columnwidth]{img/error-count-nonstrict-row--te-density-diff.pdf}
  \smallskip

  \mstrict{}
  \includegraphics[width=\columnwidth]{img/error-count-strict-row--te-density-diff.pdf}
  \caption{Changes in type error density (errors in module / lines in module) over time (seconds).}
  \label{f:error-density}
\end{figure}

@@

\section{Answers to Research Questions}
\label{s:discussion}

With the data in hand, we can answer the research questions
from~\cref{s:introduction}.
The answers confirm several aspects of Luau's design, but
at the same time offer several directions for future improvements.
Luau clearly has a powerful typechecker that can efficiently
analyze thousands of lines of code.
But, few creators use type analysis directly and the data presents little
evidence that doing so would improve their quality of life.
We conjecture that Luau needs further tailoring to match
common idioms in Roblox scripts.
Discovering these idioms is a question for future work~(\cref{s:conclusion}).

\begin{description}
  \item[How many creators choose to use type analysis?]
    A mere \pct{10} of creators opt-in to type checking.
    Most of these use \mnonstrict{} mode, and fewer than \pct{1}
    use \mstrict{} mode.

  \item[How often do projects contain modules with different analysis modes?]
    Projects rarely combine analysis modes.
    Only 512 of \nK{340} sessions ($<\pct{1}$) switched to modules with different
    analysis modes.
    This number may be a significant underapproximation because we have data
    only on modules that the creator chose to open, but the widespread use
    of \mnocheck{} mode suggests that it is not off by much.

  \item[How often do creators turn analysis off?]
    Creators rarely disable type analysis after opting in.
    Only 320 of the \nK{340} sessions ($<\pct{1}$)
    switch to a weaker mode.
    \FILL{} are they downgrading from S or NS?
    %% FILL more insight if we match upgrades + downgrades in session?

  \item[For modules that use type analysis:]\leavevmode
    \begin{description}
      \item[Which errors arise?]
        Syntax errors are extremely common~(over \pct{50},~\cref{t:type-error-count}), 
        followed by type mismatches~(\pct{20} in \mstrict{}),
        arity errors~(\pct{2}),
        and failure to unpack on optional value~(\pct{2}).
        There are several uncommon errors, such as misusing a table operation,
        and a few that never arise: internal and too-complex errors,
        duplicating or swapping a generic parameter, use of a deprecated API,
        and unsafe dynamic property lookup on a class.

      \item[How do creators respond? Which errors persist through edits?]
        Creators typically fix analysis errors.
        With few exceptions, increases and decreases in the number of errors
        balance out~(\cref{f:error-changes,t:type-error-survival,}).
        Even in \mnocheck{}, most edits that overlap with syntax errors
        remove at least one error~(\pct{67}) rather than increase or maintain
        the total.
        The errors that most-often survive edits are due to
        inference at a binary operation,
        sending \code{self} to a function that does not expect it,
        and failure to unpack an optional value.
        But the first two seldom arise, and the third is something creators
        might reasonably ignore while prototyping.

%% Keep percents
%% '((CannotInferBinaryOperation "0.50")
%%   (OptionalValueAccess "0.44")
%%   (FunctionDoesNotTakeSelf "0.40")
%%   (CannotExtendTable "0.35")
%%   (UnknownPropButGotLikeProp "0.28")
%%   (UnknownProperty "0.26")
%%   (MissingProperties "0.25")
%%   (FunctionExitsWithoutReturn "0.21")
%%   (ModuleHasCyclicDependency "0.20")
%%   (UnknownRequire "0.18")
%%   (TypeMismatch "0.18")
%%   (UnknownSymbol "0.15")
%%   (NotATable "0.12")
%%   (GenericError "0.11")
%%   (CannotCallNonFunction "0.10")
%%   (CountMismatch "0.09")
%%   (GenericExtraInformation "0.07")
%%   (IllegalRequire "0.06")
%%   (SyntaxError "0.01")

    \end{description}

  \item[What impact does type analysis have on the number of \FS{} errors?]
    Opting in to type analysis has little impact on \FS{} errors.
    The percent of \FS{} errors in \mnocheck{} code matches the percent of
    creators using \mnocheck{} mode~(\cref{f:error-by-mode}); whereas,
    if type analysis helped reduce \FS{} errors, \mnocheck{} would own a
    higher percentage of the errors.
    Furthermore, there is no apparent relation between \FS{} errors and edits,
    regardless of analysis mode~(\cref{f:error-changes}).
    A randomly-chosen edit has a 1/3 chance of increasing, decreasing, or
    maintaining the number of \FS{} errors.

\end{description}

\paragraph{Threats to Validity}
\label{s:threats}

While this study has high ecological validity due to its focus on
working creators, there are several threats to keep in mind.
Some threats stem from our method of collecting data; others
come from the limited scope of the data.

First of all, sampling is necessarily incomplete.
Though the data includes many thousands of sessions, it
may have missed a few critical sessions where a creator
adopted \mstrict{} mode, encountered \code{CodeTooComplex} errors,
and gave up.
Furthemore, our method of sampling by keystroke and by module switch
skews the data toward creators who, for whatever reason, do more
of these actions.
We have no access to other events (e.g., run locally, publish) due to privacy
concerns; future studies may wish to avoid this limitation by actively
recruiting subjects.

Second, the edit ranges are a coarse approximation of actual edits.
They begin at the lowest edited line of code and end at the highest
edited line, even if nothing between those lines changed.

Third, the total counts of type errors include a large number of syntax errors,
which have little to do with the Luau type system and are presumably easy for
creators to fix.
It would be better for the analysis to ignore these errors.
Indeed, our comparison between \FS{} errors in \mnocheck{} vs.
\mstrict{} mode~(\cref{f:error-changes}) might tell a very different story
without all the noise that syntax errors introduce.

Fourth, type analysis reports several errors at once and we have no idea which
errors, if any, creators intended to fix with their edits.
Although it appears that creators choose to fix errors,
the fixes may be a side effect of other plans.

% Telemetry cannot replace user studies, as there is no way to measure
% creator sentiment, but provide complementary data at scale.
% No idea about intent is a big problem, we don't know which error the creator wanted to fix.

Lastly, we have two remarks about external validity.
First, our information on type errors is limited to error names.
If another language has an error that is roughly similar to the
\code{CannotCallNonFunction} error but with slightly different
semantics, then the relative frequency of that error could be different
than what we observed.
Second, we know nothing about the creators in our study except that they
were selected at random from a large and diverse group.
A dataset based on a targeted subset of users might show entirely different
characteristics.

@@

\begin{table}[t]\centering
  \caption{Comparison of telemetry designs. Two orthogonal concerns are \emph{when} to collect data and \emph{whether} to add noise to enhance privacy.}
  \label{t:telemetry-design}

@@

The Blackbox project similarly collects telemetry in broad strokes.
Every Java program submitted for compilation goes into the Blackbox database (minus
any comments above the main class) along with the compiler's response~\cite{bkmu-sigcse-2014}.
%% Though somewhat intrusive,
The millions of programs in the Blackbox dataset
have enabled dozens of contributions to the area of CS
education~\cite{bask-icer-2018}, ranging from categorization of novice
errors~\cite{mk-fie-2014,m-masters-2016} to analyses of error frequency and
time-to-fix~\cite{ab-sigcse-2015}.
Related prior work used the GRUMPS telemetry system to collect 4.7 million
keystroke-level actions from learners programming in
Ada~\cite{tm-iticse-2004,tkdmceg-ascilite-2003}.
This data is not available.

@@

%\paragraph{TBD}
%
%ahmadzadeh elliman higgins iticse 2005
%
%buffardi etal 2014 adaptive and social mechanisms
%
%bddf-icse-2016
% sh edwards and jason snyder icer 2009 comparing effective and ineffective grading platform
%
%murphy etal sigcse 2009 Retina: helping students
%
%Choppella Haynes IU TR 426

@@

\FILL{} future: interviews, accommodate idioms
